---
title: "ggplot-intro-markdown"
author: "K. Walden"
date: "1/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data and Environment Setup

9. Necessary packages: `dplyr` and `ggplot2`
```{r}
# install.packages("dplyr")
# install.packages("ggplot2")
library(dplyr)
library(ggplot2)
```

10. In this tutorial, we will use the `AmesHousing` data, which provides information on the sales of individual residential properties in Ames, Iowa from 2006 to 2010. 

11. The data set contains 2930 observations, and a large number of explanatory variables involved in assessing home values. 

12. [Click here](http://jse.amstat.org/v19n3/decock/DataDocumentation.txt) for a full description of this dataset.

```{r}
# The csv file should be imported into rstudio:
AmesHousing <- read.csv("https://raw.githubusercontent.com/kwaldenphd/ggplot-intro/main/AmesHousing.csv")

str(AmesHousing)
```

# Basic Structure of `ggplot2` Functions

13. All `ggplot` functions must have at least three components:
- ***data frame***: In this activity we will be using the AmesHousing data.
- ***geom***: to determine the type of geometric shape used to display the data, such as line, bar, point, or area.
- ***aes***: to determine how variables in the data are mapped to visual properties (aesthetics) of geoms. This can include x position, y position, color, shape, fill, and size.

14. Thus the simplest code for a graphic made with `ggplot()` would have one of the the following forms:
- `ggplot(data, aes(x, y)) + geom_line()`
or
- `ggplot(data) + geom_line(aes(x, y))`

15. These two lines of code provide identical results. 

16. In the first case, the `aes` is set for all `geoms`, meaning the same x and y variables are mapped to any `geoms` that are added. 

17. Because more complex graphics can include multiple `geoms`, it can be advantageous to locally define the `aes` for each `geom` as shown in the second line of code.

```{r}
# Create a histogram of housing prices
ggplot(data=AmesHousing) + geom_histogram(mapping = aes(x = SalePrice))
```

18. In the above code, the terms `data=`, `mapping=`, and `x=` are not required, but are used for clarification. 

19. For example, the following code will produce identical results: 

```{r}
ggplot(AmesHousing) + geom_histogram(aes(SalePrice))
```

```{r}
# Create a scatterplot of above ground living area by sales price
ggplot(data=AmesHousing) + geom_point(mapping= aes(x=GrLivArea, y=SalePrice))
```

Q1: Open your R Script and add a comment beneath your earlier answers indicating all subsequent work pertains to the ggplot lab. Beneath this comment, write code that does the following:
 -Creates the same histogram as the example above, but modifies the code so that the `aes` is not within the `geom`
- Creates a scatterplot using `ggplot` with `Fireplaces` as the x-axis and `SalePrice` as the y-axis

# Customizing Graphics Using `ggplot2`

20. In this section, we layer additional components onto the two graphs shown above.

```{r}
ggplot(data=AmesHousing) +                         
      geom_histogram(mapping = aes(SalePrice/100000), 
          breaks=seq(0, 7, by = 1), col="red", fill="lightblue") + 
      geom_density(mapping = aes(x=SalePrice/100000, y = (..count..)))  +   
      labs(title="Figure 9: Housing Prices in Ames, Iowa (in $100,000)", 
          x="Sale Price of Individual Homes")   
```

21. The `histogram` `geom` (`geom_histogram`) transforms the SalePrice, modifies the bin size and changes the color.

22. `geom_density` overlays a density curve on top of the histogram.

23. Typically density curves and histrograms have very different scales, here we use `y = (..count..)` to modify the density. Alternatively, we could specify `aes(x = SalePrice/100000, y = (..density..))` in the `histogram` `geom`.

24. The `labs()` command adds a title and an x-axis label. 

25. A y-axis label can also be added by using `y = " `.

26. In the code below we create three scatterplots of the log of the above ground living area by the log of sales price.

```{r}
ggplot(data=AmesHousing, aes(x=log(GrLivArea), y=log(SalePrice)) ) +      
  geom_point(shape = 3, color = "darkgreen") +                                     
  geom_smooth(method=lm,  color="green") +                  
  labs(title="Figure 10: Housing Prices in Ames, Iowa")
```

```{r}
ggplot(data=AmesHousing) + 
  geom_point(aes(x=log(GrLivArea), y=log(SalePrice), color=KitchenQual),shape=2, size=2) + 
  geom_smooth(aes(x=log(GrLivArea), y=log(SalePrice), color=KitchenQual), 
          method=loess, size=1) +                        
  labs(title="Figure 11: Housing Prices in Ames, Iowa") 
```

```{r}
ggplot(data=AmesHousing) +
  geom_point(mapping = aes(x=log(GrLivArea), y=log(SalePrice), color=KitchenQual)) +
  geom_smooth(mapping = aes(x=log(GrLivArea), y=log(SalePrice), color=KitchenQual), 
      method=lm, se=FALSE, fullrange=TRUE) +                             
  facet_grid( ~ Fireplaces) +                      
  labs(title="Figure 12: Housing Prices in Ames, Iowa")
```

27. `geom_point` is used to create a scatterplot.

28. Multiple shapes can be used as points. The [Data Visualization Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) lists several shape options

29. `geom_smooth` adds a fitted line through the data.
- `method=lm` specifies a linear regression line. method=loess creates a smooth fit curve
- `se=FALSE` removes the shaded confidence regions around each line
- `fullrange=TRUE` extends all regression lines to the same length

30. `facet_grid` and `facet_wrap` commands are used to create multiple plots. 

31. In these examples, we have created separate scatterplots based upon the number of fireplaces.

32. When assigning fixed characteristics, (such as color, shape or size), the commands occur outside the `aes`, as in `color="green"`. 

33. When characteristics are dependent on the data, the command should occur within the `aes`, as in `color=Kitchen.Qual`.

34. In the above examples, only a few `geoms` are listed. 

35. The [`ggplot2` website](https://ggplot2.tidyverse.org/reference/) lists each `geom` and gives detailed examples of how they are used.

Q2: Add a comment indicating you’re answering question 2, beneath the comment write code that does the following:
- Creates a scatterplot using `YearBuilt` as the explanatory variable and `SalePrice` as the response variable. 
- Include a regression line, a title, and labels for the x and y axes.
- Modifies the scatterplot by replacing the regression line with a loess smoother
- Colors the points by the overall condition of the home, `OverallCond`
 
# Additional Considerations

## Data Types

36. It's important to think about how the type of data you are working with shapes choices you make around visualization.

37. If you use the `str` command after reading data into `R`, you will notice that each variable is assigned one of the following types: 
- `Character`
- `Numeric` (real numbers)
- `Integer`
- `Complex`
- `Logical` (TRUE/FALSE)

38. In particular, the variable `Fireplaces` in considered an integer. 

39. In the code below we try to color and fill a density graph by an integer value. 

40. Notice that the `color` and `fill` commands appear to be ignored in the graph.

```{r}
# str(AmesHousing)
ggplot(data=AmesHousing) +                   
  geom_density(aes(SalePrice, color = Fireplaces,  fill = Fireplaces))
```

41. In the following code, we use the `dplyr` package to modify the AmesHousing data.

42. We first restrict the dataset to only houses with less than three fireplaces and then create a new variable, called `Fireplace2`. 

43. The `as.factor` command creates a factor, which is a variable that contains a set of numeric codes that map to character-valued levels. 

44. Notice that the `color` and `fill` commands now work properly.

```{r}
# Create a new data frame with only houses with less than 3 fireplaces
AmesHousing2 <- filter(AmesHousing, Fireplaces < 3)

# Create a new variable called Fireplace2
AmesHousing2 <-mutate(AmesHousing2,Fireplace2=as.factor(Fireplaces))

str(AmesHousing2)

ggplot(data=AmesHousing2) +                 
  geom_density(aes(SalePrice, color = Fireplace2,  fill = Fireplace2), alpha = 0.2)
```

## Customizing Graphs

45. In addition to using a `data.frame`, `geoms`, and `aes`, several additional components can be added to customize each graph:
- stats
- scales
- themes
- positions
- coordinate systems
- labels
- legends

46. We will not discuss all of these components here, but the materials in the [Additional Resources](#additional-resources) section provide detailed explanations. 

47. In the code below we provide a few examples on how to customize graphs.

```{r}
ggplot(AmesHousing2, aes(x = Fireplace2, y = SalePrice, color = PavedDrive)) +
  geom_boxplot(position = position_dodge(width = 1)) +
  coord_flip()+ 
  labs(title="Housing Prices in Ames, Iowa") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "blue", face="bold", size=12, hjust=0))
```

48. `position` is used to address `geoms` that would take the same space on a graph. 

49. In the above boxplot, `position_dodge(width = 1)` adds a space between each box. 

50. For scatterplots, `position = position_jitter()` puts spaces between overlapping points.

51. `theme` is used to change the style of a graph, but does not change the data or geoms. 

52. The above code is used to modify only the title in a boxplot. 

53. A better approach for beginners is to choose among themes that were created to customize the overall graph. 

54. Common examples include:
- `theme_bw()`
- `theme_classic()`
- `theme_grey()`
- `theme_minimal()`

55. You can also install the [`ggthemes` package](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) for many more options.

# Additional Resources

56. Additional `ggplot2` resources:
- Roger Peng, "Plotting with ggplot2" *YouTube* 
  * [Part 1](https://youtu.be/HeqHMM4ziXA) (14 October 2013)
  * [Part 2](https://youtu.be/n8kYa9vu1l8) (14 October 2013)
- [RStudio "Data Visualization With `ggplot2`" cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [`ggplot2` overview](https://ggplot2.tidyverse.org/), Tidyverse documentation
- [`ggplot2` package documentation](http://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf), R project
- Basel Institute for Clinical Epidemiology and Biostatistics, [`ggplot2` tutorial](http://bbs.ceb-institute.org/wp-content/uploads/2011/09/handout_ggplot2.pdf)
- [Questions tagged `ggplot2` on Stackoverflow](https://stackoverflow.com/tags/ggplot2)
- Winston Chang, [*R Graphics Cookbook: Practical Recipes for Visualizing Data*](http://www.cookbook-r.com/Graphs/), O'Reilly (2013).
- Hadley Wickham, Danielle Navarro, and Thomas Lin Pedersen, [*ggplot2: elegant graphics for data analysis*](https://ggplot2-book.org/), Springer (in-progress 3rd edition).

# Additional Questions

## Question 3

```{r}
ggplot(AmesHousing2, aes(x = Fireplace2, y = SalePrice, color = PavedDrive)) +
  geom_boxplot(position = position_dodge(width = 1)) +
  coord_flip()+ 
  labs(title="Housing Prices in Ames, Iowa") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "blue", face="bold", size=12, hjust=0))
```

Add a comment indicating you’re answering question 3, beneath the comment answer the following:
- In the density plot above, explain what the `color`, `fill`, and `alpha` commands are used for. 
  * Hint: try running the code with and without these commands or use the Data Visualization Cheat Sheet.
- In the boxplot, what is done by the code `coord_flip()`?
- Create a new boxplot, similar to the one above, but use `theme_bw()` instead of the given theme command. Explain how this changes the graph.

## Question 4

Add a comment to your lab write-up indicating you’re working on question 4. This question requires you to use the `dplyr` package to manipulate the dataset before making any graphics.
- Restrict the AmesHousing data to only sales under normal conditions. 
  * In other words, `SaleCondition == "Norm"`
- Create a new variable called `TotalSqFt = GRLivArea  +  TotalBsmtSF` and remove any homes with more than 3000 total square feet.
- With this modified dataset, create a graphic involving no more than three explanatory variables that best illustrates how to predict home’s sale price. 

# Lab Notebook Questions

Q1: Open your R Script and add a comment beneath your earlier answers indicating all subsequent work pertains to the ggplot lab. Beneath this comment, write code that does the following:
 -Creates the same histogram as the example above, but modifies the code so that the `aes` is not within the `geom`
- Creates a scatterplot using `ggplot` with `Fireplaces` as the x-axis and `SalePrice` as the y-axis

Q2: Add a comment indicating you’re answering question 2, beneath the comment write code that does the following:
- Creates a scatterplot using `YearBuilt` as the explanatory variable and `SalePrice` as the response variable. 
- Include a regression line, a title, and labels for the x and y axes.
- Modifies the scatterplot by replacing the regression line with a loess smoother
- Colors the points by the overall condition of the home, `OverallCond`

```{r}
ggplot(AmesHousing2, aes(x = Fireplace2, y = SalePrice, color = PavedDrive)) +
  geom_boxplot(position = position_dodge(width = 1)) +
  coord_flip()+ 
  labs(title="Housing Prices in Ames, Iowa") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "blue", face="bold", size=12, hjust=0))
```

Q3: Add a comment indicating you’re answering question 3, beneath the comment answer the following:
- In the density plot above, explain what the `color`, `fill`, and `alpha` commands are used for. 
  * Hint: try running the code with and without these commands or use the Data Visualization Cheat Sheet.
- In the boxplot, what is done by the code `coord_flip()`?
- Create a new boxplot, similar to the one above, but use `theme_bw()` instead of the given theme command. Explain how this changes the graph

Q4: Add a comment to your lab write-up indicating you’re working on question 4. This question requires you to use the `dplyr` package to manipulate the dataset before making any graphics.
- Restrict the AmesHousing data to only sales under normal conditions. 
  * In other words, `SaleCondition == "Norm"`
- Create a new variable called `TotalSqFt = GRLivArea  +  TotalBsmtSF` and remove any homes with more than 3000 total square feet.
- With this modified dataset, create a graphic involving no more than three explanatory variables that best illustrates how to predict home’s sale price